Component response rate variation drives stability in large complex systems

### 2018-05-18 ms.Rmd
More writing

**The stability of a complex system generally decreases with increasing system size, as is demonstrated by random matrix theory [@May1972; @Allesina2012]. This counter-intuitive result, first shown by May [@May1972], is broadly relevant for understanding the dynamics and persistence of systems such as ecological [@May1972; @Allesina2012], neurological [@Gray2008; @Gray2009], biochemical [@Rosenfeld2009; @MacArthur2010] and socio-economic [@Haldane2011; @Suweis2014; @Bardoscia2017] networks. Much attention has especially been given to the stability of ecological communities such as food webs or mutualist networks, with recent work investigating how different community structures affect stability [@Allesina2012; @Mougi2012; @Allesina2015a; @Gao2016; @Grilli2017; @Patel2018]. But more broadly, stabilising mechanisms in complex systems remain under-developed, and the effect of variation in the response rate of individual system components remains an open problem [@Allesina2015]. Here I show that when components of a complex system respond to system dynamics at different rates ($\boldsymbol{\gamma}$), system stability is markedly increased. Stability increases despite higher variation among interaction strengths ($\boldsymbol{\sigma^{2}}$) due to the clustering of eigenvalues at the centre of eigenvalue distributions. This effect of variation in $\boldsymbol{\gamma}$ becomes increasingly important as system size increases, to the extent that the largest stable complex systems would otherwise be unstable if not for $\boldsymbol{Var(\gamma)}$. My results therefore reveal a previously unconsidered driver of system stability that is likely to be pervasive across all complex systems. Future research in complex systems should therefore account for the varying response rates of individual system components when assessing whole system stability.**

@@

@@

>>>>

**The stability of a complex system generally decreases with increasing system size, as is demonstrated by random matrix theory [@May1972; @Allesina2012]. This counter-intuitive result, first shown by May [@May1972], is broadly relevant for understanding the dynamics and persistence of systems such as ecological [@May1972; @Allesina2012], neurological [@Gray2008; @Gray2009], biochemical [@Rosenfeld2009; @MacArthur2010] and socio-economic [@Haldane2011; @Suweis2014; @Bardoscia2017] networks. Much attention has especially been given to the stability of ecological communities such as food webs or mutualist networks, with recent work investigating how different community structures affect stability [@Allesina2012; @Mougi2012; @Allesina2015a; @Gao2016; @Grilli2017; @Patel2018]. But more broadly, stabilising mechanisms in complex systems remain under-developed, and the effect of variation in the response rate of individual system components remains an open problem [@Allesina2015]. Here I show that when components of a complex system respond to system dynamics at different rates ($\boldsymbol{\gamma}$), system stability is markedly increased. Stability increases due to the clustering of some eigenvalues toward the centre of eigenvalue distributions despite the destabilising effect of higher variation among interaction strengths ($\boldsymbol{\sigma^{2}}$). This effect of variation in $\boldsymbol{\gamma}$ becomes increasingly important as system size increases, to the extent that the largest stable complex systems would otherwise be unstable if not for $\boldsymbol{Var(\gamma)}$. My results therefore reveal a previously unconsidered driver of system stability that is likely to be pervasive across all complex systems. Future research in complex systems should therefore account for the varying response rates of individual system components when assessing whole system stability.**

@@

Randomly assembled complex systems can be represented as large square matrices ($M$) with $S$ components (e.g., species[@Allesina2012] or banks[@Haldane2011]). One element of such a matrix $M_{ij}$ defines how component $j$ affects component $i$ in the system at a point of equilibrium[@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation (e.g., carrying capacity in ecological communities). Traditionally, values of off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma^{2}$; diagonal elements are set to -1[@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $M$ are negative ($\max\left(\Re(\lambda)\right) < 0$). In a large system (high $S$), eigenvalues are distributed uniformly[@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma\sqrt{SC}$[@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma^{2}$ increase.

The above stability criteria assumes that individual components respond to perturbations of the system at the same rate ($\gamma$), but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perterbation will depend on the generation time of individuals, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individual institutions respond to perturbations in complex social-ecological systems, is likely to vary. The effect that this variance in the rate at which individual system components respond to perturbations has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, complex systems in which $\gamma$ varies actually tend to be more stable, especially when $S$ is high.

```{r echo = FALSE}
S     <- 200;
C     <- 0.05
sigma <- 0.4;
```

Rows in $M$ define how a given component $i$ is affected by other components of the system, meaning that the rate of component response time can be modelled by multiplying all row elements by a scalar value $\gamma_{i}$[@Patel2018]. The distribution of $\gamma$ over $S$ components thereby models the distribution of component response rates. An instructive example compares one $M$ where $\gamma_{i} = 1$ for all $i$ in $S$ to the same $M$ when half of $\gamma_{i} = 1.95$ and half of $\gamma_{i} = 0.05$. This models one system in which $\gamma$ is invariant and one in which $\gamma$ varies, but systems are otherwise identical (note $E[\gamma_{i}] = 1$ in both cases). I assume $S = 200$, $C = 0.05$, $\sigma = 0.4$; diagonal elements are set to $-1$ and non-zero off-diagonal elements are drawn from $\mathcal{N}(0, \sigma^{2})$. Eigenvalues of $M$ are distributed uniformly within a circle centred at ($0, -1$) with a radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig. 1a). Hence, the real components of eigenvalues are highly unlikely to all be negative when all $\gamma_{i} = 1$. But when

<!--- Say that the amnat paper did this with multiplying a row --->

@@

===== ACTUALLY, use paragraph three to shift to considering component time scales not having been done (skip 3, sort of).
